---
title: "04 — Hot/Cold × p16 Overlap Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

Core analysis notebook: tests the relationship between p16/senescence status and
hot/cold immune phenotype.

**Hypotheses tested:**

1. SASP-high samples → hot (pro-inflammatory cytokines drive TIL infiltration)
2. CDKN2A-deleted samples → cold (loss of senescence → loss of SASP signaling)
3. Relationship is tumor-type specific

```{r setup, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(ggplot2)
library(rstatix)
library(pheatmap)

set.seed(42)

DATA    <- "../data"
FIGURES <- normalizePath("../figures", mustWork = FALSE)
dir.create(FIGURES, showWarnings = FALSE)
```

## 1. Load & Merge Data

```{r load-data}
stopifnot(
  file.exists(file.path(DATA, "sample_immune_labels.parquet")),
  file.exists(file.path(DATA, "sample_senescence.parquet")),
  file.exists(file.path(DATA, "tcga_clinical_survival.tsv"))
)

immune <- read_parquet(file.path(DATA, "sample_immune_labels.parquet")) |>
  rename(sample_id = `__index_level_0__`)
senes  <- read_parquet(file.path(DATA, "sample_senescence.parquet")) |>
  rename(sample_id = `__index_level_0__`)

clin <- read.table(
  file.path(DATA, "tcga_clinical_survival.tsv"),
  sep = "\t", header = TRUE, row.names = 1, stringsAsFactors = FALSE
)

master <- inner_join(immune, senes, by = "sample_id") |>
  left_join(
    clin[, "age_at_initial_pathologic_diagnosis", drop = FALSE] |>
      tibble::rownames_to_column("sample_id"),
    by = "sample_id"
  ) |>
  rename(age = age_at_initial_pathologic_diagnosis)

cat("Master table:", nrow(master), "rows,", ncol(master), "cols\n")
cat("Columns:", paste(colnames(master), collapse = ", "), "\n")

# Tumor types with ≥100 samples — used for confirmatory primary tests
# (pan-cancer analyses still use all types)
confirmatory_types <- master |>
  count(tumor_type) |>
  filter(n >= 100) |>
  pull(tumor_type)
message(sprintf("Confirmatory set: %d tumor types with ≥100 samples (of %d total)",
                length(confirmatory_types), n_distinct(master$tumor_type)))
master_confirmatory <- master |> filter(tumor_type %in% confirmatory_types)
```

> **Pre-specified primary hypotheses**
> H1: ρ(CDKN2A_expr, IFN-γ) > 0 — tested in §2 below
> H2: SASP higher in hot tumors — tested in §5
> H3: CDKN2A deletion enriched in cold — tested in §4
> H4: p16-high hot tumors best OS — tested in notebook 05 §4
> H5: CDKN2A_expr independent OS predictor (age-adjusted) — notebook 05 §6
> All other analyses in this notebook are **exploratory**.

```{r testing-strategy}
# Primary hypotheses H1-H5: Bonferroni-corrected threshold (5 tests, α=0.05)
ALPHA_PRIMARY    <- 0.05 / 5   # = 0.01
# Exploratory analyses: BH FDR, reported separately
ALPHA_EXPLORATORY <- 0.05
message(sprintf("Primary α = %.3f (Bonferroni, 5 pre-specified hypotheses)", ALPHA_PRIMARY))
message(sprintf("Exploratory α = %.2f (BH FDR)", ALPHA_EXPLORATORY))
```

## 2. CDKN2A Expression vs IFN-γ Score (Scatter)

```{r scatter-pancancer}
scatter_data <- master |>
  select(CDKN2A_expr, ifng_score, tumor_type) |>
  filter(!is.na(CDKN2A_expr), !is.na(ifng_score))

# PRIMARY TEST (H1): compare p to ALPHA_PRIMARY
# Using master_confirmatory (n≥100 per type) for primary hypothesis test
scatter_confirmatory <- master_confirmatory |>
  select(CDKN2A_expr, ifng_score, tumor_type) |>
  filter(!is.na(CDKN2A_expr), !is.na(ifng_score))
ct_confirmatory <- cor.test(scatter_confirmatory$CDKN2A_expr, scatter_confirmatory$ifng_score,
                            method = "spearman", exact = FALSE)
cat(sprintf("Pan-cancer Spearman (confirmatory set): rho=%.3f, p=%.2e\n",
            ct_confirmatory$estimate, ct_confirmatory$p.value))

ct <- cor.test(scatter_data$CDKN2A_expr, scatter_data$ifng_score,
               method = "spearman", exact = FALSE)
cat(sprintf("Pan-cancer Spearman (full cohort): rho=%.3f, p=%.2e\n", ct$estimate, ct$p.value))

p <- ggplot(scatter_data, aes(x = CDKN2A_expr, y = ifng_score)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "steelblue") +
  labs(
    x     = "CDKN2A expression (log2)",
    y     = "IFN-\u03b3 signature score",
    title = sprintf("CDKN2A vs IFN-\u03b3 score (\u03c1=%.2f, p=%.1e)",
                    ct$estimate, ct$p.value)
  ) +
  theme_bw()

ggsave(file.path(FIGURES, "cdkn2a_vs_ifng_scatter.png"),
       plot = p, width = 7, height = 6, dpi = 150)
p
```

## 3. Spearman Correlations: Senescence Scores × Immune Scores

### Pan-cancer correlation table

```{r spearman-pancancer}
senes_cols  <- intersect(
  c("CDKN2A_expr", "SASP", "Senescence_Effectors",
    "senescence_up", "reactome_senescence",
    "senmayo", "hallmark_p53", "gobp_senescence", "senescence_dn",
    "CDKN1A_expr", "SERPINE1_expr", "GDF15_expr", "LMNB1_expr"),
  colnames(master)
)
immune_cols <- intersect(c("ifng_score", "cd8_fraction"), colnames(master))

# Build tidy correlation table with BH FDR correction
cor_long <- master |>
  select(all_of(c(senes_cols, immune_cols))) |>
  cor_test(vars  = all_of(senes_cols),
           vars2 = all_of(immune_cols),
           method = "spearman") |>
  adjust_pvalue(method = "BH") |>
  add_significance()

print(cor_long)

# Wide rho matrix for heatmap
rho_mat <- cor_long |>
  select(var1, var2, cor) |>
  tidyr::pivot_wider(names_from = var2, values_from = cor) |>
  tibble::column_to_rownames("var1") |>
  as.matrix()

png(file.path(FIGURES, "spearman_heatmap_pancancer.png"),
    width = 700, height = 500, res = 150)
pheatmap(rho_mat,
         display_numbers = TRUE,
         number_format   = "%.2f",
         color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
         breaks           = seq(-0.6, 0.6, length.out = 101),
         main             = "Spearman \u03c1: Senescence vs Immune (pan-cancer)",
         fontsize         = 11)
dev.off()

pheatmap(rho_mat,
         display_numbers = TRUE,
         number_format   = "%.2f",
         color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
         breaks           = seq(-0.6, 0.6, length.out = 101),
         main             = "Spearman \u03c1: Senescence vs Immune (pan-cancer)",
         fontsize         = 11)
```

### Per-tumor-type CDKN2A vs IFN-γ

```{r spearman-per-type}
# EXPLORATORY: interpret with BH FDR, not primary threshold
if ("tumor_type" %in% colnames(master)) {
  per_type <- master |>
    filter(!is.na(CDKN2A_expr), !is.na(ifng_score)) |>
    group_by(tumor_type) |>
    # Minimum n=30 per group for stable Spearman estimates
    filter(n() >= 30) |>
    group_modify(\(dat, grp) {
      ct <- cor.test(dat$CDKN2A_expr, dat$ifng_score, method = "spearman", exact = FALSE)
      tibble(rho = ct$estimate, pval = ct$p.value, n = nrow(dat))
    }) |>
    ungroup() |>
    mutate(
      padj  = p.adjust(pval, method = "BH"),
      sig   = padj < 0.05
    ) |>
    arrange(desc(rho))

  print(per_type)

  p2 <- ggplot(per_type, aes(x = reorder(tumor_type, rho),
                              y = rho, fill = rho > 0)) +
    geom_col() +
    scale_fill_manual(values = c("TRUE" = "#e74c3c", "FALSE" = "#3498db"),
                      guide = "none") +
    geom_hline(yintercept = 0, linewidth = 0.5) +
    coord_flip() +
    labs(
      x     = NULL,
      y     = "Spearman \u03c1 (CDKN2A vs IFN-\u03b3)",
      title = "Per-tumor-type correlation: CDKN2A expression vs immune hotness"
    ) +
    theme_bw()

  ggsave(file.path(FIGURES, "cdkn2a_ifng_corr_by_type.png"),
         plot = p2, width = 8, height = 10, dpi = 150)
  p2
}
```

## 4. CDKN2A Deletion vs Hot/Cold (Chi-squared)

```{r chi-squared}
# PRIMARY TEST (H3): compare p to ALPHA_PRIMARY
if ("cdkn2a_deleted" %in% colnames(master) && "hot_cold" %in% colnames(master)) {
  chi_data <- master |> filter(!is.na(cdkn2a_deleted), !is.na(hot_cold))
  ct_tbl <- table(
    deletion = ifelse(chi_data$cdkn2a_deleted, "CDKN2A deleted", "CDKN2A intact"),
    immune   = chi_data$hot_cold
  )
  print(ct_tbl)

  chi_res <- chisq.test(ct_tbl)
  if (any(chi_res$expected < 5))
    warning("Chi-squared (full cohort): expected cell counts < 5 — results may be unreliable")
  cat(sprintf("\nChi-squared test (full cohort): chi2=%.2f, p=%.2e, df=%d\n",
              chi_res$statistic, chi_res$p.value, chi_res$parameter))

  # Using master_confirmatory (n≥100 per type) for primary hypothesis test
  chi_data_conf <- master_confirmatory |> filter(!is.na(cdkn2a_deleted), !is.na(hot_cold))
  ct_tbl_conf <- table(
    deletion = ifelse(chi_data_conf$cdkn2a_deleted, "CDKN2A deleted", "CDKN2A intact"),
    immune   = chi_data_conf$hot_cold
  )
  chi_res_conf <- chisq.test(ct_tbl_conf)
  if (any(chi_res_conf$expected < 5))
    warning("Chi-squared (confirmatory): expected cell counts < 5 — results may be unreliable")
  cat(sprintf("Chi-squared test (confirmatory set): chi2=%.2f, p=%.2e, df=%d\n",
              chi_res_conf$statistic, chi_res_conf$p.value, chi_res_conf$parameter))

  # Hot fraction by deletion status
  hot_frac <- master |>
    group_by(cdkn2a_deleted) |>
    summarise(hot_fraction = mean(hot_cold == "hot", na.rm = TRUE), .groups = "drop")
  print(hot_frac)
}
```

## 5. SASP Score Distribution: Hot vs Cold (Mann-Whitney U)

```{r mann-whitney}
# PRIMARY TEST (H2): compare p to ALPHA_PRIMARY
if ("SASP" %in% colnames(master) && "hot_cold" %in% colnames(master)) {
  hot_sasp  <- master |> filter(hot_cold == "hot",  !is.na(SASP)) |> pull(SASP)
  cold_sasp <- master |> filter(hot_cold == "cold", !is.na(SASP)) |> pull(SASP)

  mw <- wilcox.test(hot_sasp, cold_sasp, alternative = "two.sided", exact = FALSE)
  cat(sprintf("Mann-Whitney U: W=%.0f, p=%.2e\n", mw$statistic, mw$p.value))
  cat(sprintf("Median SASP — hot: %.3f,  cold: %.3f\n",
              median(hot_sasp), median(cold_sasp)))

  plot_df <- master |>
    filter(!is.na(SASP), !is.na(hot_cold)) |>
    select(SASP, hot_cold)

  p3 <- ggplot(plot_df, aes(x = SASP, fill = hot_cold)) +
    geom_histogram(bins = 60, alpha = 0.6, position = "identity") +
    scale_fill_manual(values = c("hot" = "#e74c3c", "cold" = "#3498db")) +
    labs(
      x     = "SASP ssGSEA score",
      title = sprintf("SASP score distribution: hot vs cold tumors\n(p=%.2e)", mw$p.value),
      fill  = NULL
    ) +
    theme_bw()

  ggsave(file.path(FIGURES, "sasp_score_hot_cold.png"),
         plot = p3, width = 7, height = 5, dpi = 150)
  p3
}
```

## 6. New Expression Marker Correlations vs Immune Scores

```{r new-marker-correlations}
new_expr_cols <- intersect(
  c("CDKN1A_expr", "SERPINE1_expr", "GDF15_expr", "LMNB1_expr"),
  colnames(master)
)
immune_cols <- intersect(c("ifng_score", "cd8_fraction"), colnames(master))

if (length(new_expr_cols) > 0 && length(immune_cols) > 0) {
  new_cor <- master |>
    select(all_of(c(new_expr_cols, immune_cols))) |>
    cor_test(vars  = all_of(new_expr_cols),
             vars2 = all_of(immune_cols),
             method = "spearman") |>
    adjust_pvalue(method = "BH") |>
    add_significance()

  print(new_cor)

  # Heatmap of new markers
  new_rho_mat <- new_cor |>
    select(var1, var2, cor) |>
    tidyr::pivot_wider(names_from = var2, values_from = cor) |>
    tibble::column_to_rownames("var1") |>
    as.matrix()

  png(file.path(FIGURES, "spearman_heatmap_new_markers.png"),
      width = 700, height = 500, res = 150)
  pheatmap(new_rho_mat,
           display_numbers = TRUE,
           number_format   = "%.2f",
           color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
           breaks           = seq(-0.6, 0.6, length.out = 101),
           main             = "Spearman \u03c1: New Senescence Markers vs Immune (pan-cancer)",
           fontsize         = 11)
  dev.off()

  pheatmap(new_rho_mat,
           display_numbers = TRUE,
           number_format   = "%.2f",
           color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
           breaks           = seq(-0.6, 0.6, length.out = 101),
           main             = "Spearman \u03c1: New Senescence Markers vs Immune (pan-cancer)",
           fontsize         = 11)
} else {
  cat("New expression marker columns not found — re-run notebook 03 first.\n")
}
```

## 7. Senescence Gene Alteration Landscape — Hot vs Cold

```{r genomic-alterations}
library(purrr)

# Map display label → column in master
# Prefer _altered (del|mut) where available, fall back to _deleted or _mutated
alt_gene_map <- list(
  CDKN2A = "cdkn2a_altered",
  TP53   = "tp53_altered",
  RB1    = "rb1_altered",
  ATM    = "atm_altered",
  PTEN   = "pten_altered",
  CDKN1A = "cdkn1a_deleted",
  CDKN2B = "cdkn2b_deleted",
  CDKN2C = "cdkn2c_deleted"
)

# Build tidy summary: % altered in hot vs cold per gene, with chi-squared p
alt_summary <- map_dfr(names(alt_gene_map), function(gene) {
  col <- alt_gene_map[[gene]]
  if (!col %in% colnames(master)) return(NULL)

  sub <- master |> filter(!is.na(.data[[col]]), !is.na(hot_cold))
  if (nrow(sub) < 10) return(NULL)

  ct_tbl <- table(immune = sub$hot_cold,
                  altered = sub[[col]])
  chi_p <- tryCatch({
    cr <- chisq.test(ct_tbl)
    if (any(cr$expected < 5))
      warning(sprintf("Chi-squared (%s): expected cell counts < 5 — results may be unreliable", gene))
    cr$p.value
  }, error = function(e) NA_real_)

  sub |>
    group_by(hot_cold) |>
    summarise(
      gene        = gene,
      pct_altered = 100 * mean(.data[[col]], na.rm = TRUE),
      n           = n(),
      .groups     = "drop"
    ) |>
    mutate(chi_p = chi_p)
})

if (nrow(alt_summary) > 0) {
  print(alt_summary |> select(gene, hot_cold, pct_altered, n, chi_p) |> arrange(gene))

  # ── Grouped bar chart ─────────────────────────────────────────────────────
  # Order genes by overall alteration rate descending
  gene_order <- alt_summary |>
    group_by(gene) |>
    summarise(mean_pct = mean(pct_altered), .groups = "drop") |>
    arrange(desc(mean_pct)) |>
    pull(gene)

  # Chi-squared significance label for annotation
  chi_labels <- alt_summary |>
    distinct(gene, chi_p) |>
    mutate(
      sig = case_when(
        chi_p < ALPHA_PRIMARY      ~ "**",   # primary threshold (Bonferroni-corrected)
        chi_p < ALPHA_EXPLORATORY  ~ "*",    # exploratory threshold
        TRUE                       ~ ""
      )
    )

  p7 <- ggplot(
    alt_summary |> mutate(gene = factor(gene, levels = gene_order)),
    aes(x = gene, y = pct_altered, fill = hot_cold)
  ) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(
      data = chi_labels |> mutate(gene = factor(gene, levels = gene_order)),
      aes(x = gene, y = -1.5, label = sig),
      inherit.aes = FALSE, size = 4, fontface = "bold", colour = "grey30"
    ) +
    scale_fill_manual(
      values = c("hot" = "#e74c3c", "cold" = "#3498db"),
      name   = "Immune\nphenotype"
    ) +
    labs(
      x     = NULL,
      y     = "% of samples altered",
      title = "Senescence gene alteration rates: hot vs cold tumors",
      caption = "Alteration = deletion and/or mutation; * p<0.05 (exploratory), ** p<0.01 (primary threshold) (chi-squared)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, face = "bold"))

  ggsave(file.path(FIGURES, "alteration_hot_cold_grouped_bar.png"),
         plot = p7, width = 9, height = 5, dpi = 150)
  p7

  # ── Heatmap: % altered per gene × tumor type, split by hot/cold ──────────
  # Compute per-type rates for hot tumors only (cold shown as facet or separate)
  type_alt <- map_dfr(names(alt_gene_map), function(gene) {
    col <- alt_gene_map[[gene]]
    if (!col %in% colnames(master)) return(NULL)
    master |>
      filter(!is.na(.data[[col]]), !is.na(hot_cold), !is.na(tumor_type)) |>
      group_by(tumor_type, hot_cold) |>
      summarise(gene = gene,
                pct_altered = 100 * mean(.data[[col]], na.rm = TRUE),
                n = n(), .groups = "drop")
  })

  if (nrow(type_alt) > 0) {
    p7b <- ggplot(
      type_alt |> mutate(gene = factor(gene, levels = gene_order)),
      aes(x = gene, y = tumor_type, fill = pct_altered)
    ) +
      facet_wrap(~ hot_cold, ncol = 2) +
      geom_tile(colour = "white", linewidth = 0.3) +
      scale_fill_gradient(low = "white", high = "#c0392b",
                          name = "% altered", limits = c(0, NA)) +
      labs(
        x     = NULL,
        y     = NULL,
        title = "Senescence gene alteration rate by tumor type and immune phenotype"
      ) +
      theme_bw() +
      theme(
        axis.text.x  = element_text(angle = 35, hjust = 1, face = "bold"),
        strip.text   = element_text(face = "bold", size = 11)
      )

    ggsave(file.path(FIGURES, "alteration_hot_cold_heatmap.png"),
           plot = p7b, width = 14, height = 10, dpi = 150)
    p7b
  }
} else {
  cat("No alteration columns found — re-run notebooks 01 and 03 first.\n")
}
```

## 8. Age & Senescence Diagnostics

### 8a. Age by tumor type

```{r age-by-tumor-type}
if ("age" %in% colnames(master) && "tumor_type" %in% colnames(master)) {
  age_type <- master |>
    filter(!is.na(age), !is.na(tumor_type)) |>
    group_by(tumor_type) |>
    mutate(median_age = median(age, na.rm = TRUE)) |>
    ungroup() |>
    mutate(tumor_type = reorder(tumor_type, median_age))

  p8a <- ggplot(age_type, aes(x = tumor_type, y = age)) +
    geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.size = 0.5) +
    coord_flip() +
    labs(
      x     = NULL,
      y     = "Age at diagnosis (years)",
      title = "Patient age distribution by tumor type"
    ) +
    theme_bw()

  ggsave(file.path(FIGURES, "age_by_tumor_type.png"),
         plot = p8a, width = 8, height = 10, dpi = 150)
  p8a
} else {
  cat("age or tumor_type column not found.\n")
}
```

### 8b. Age by hot/cold immune phenotype

```{r age-hot-cold}
if ("age" %in% colnames(master) && "hot_cold" %in% colnames(master)) {
  age_hc <- master |> filter(!is.na(age), !is.na(hot_cold))

  mw_age <- wilcox.test(age ~ hot_cold, data = age_hc, exact = FALSE)
  med_hot  <- median(age_hc$age[age_hc$hot_cold == "hot"],  na.rm = TRUE)
  med_cold <- median(age_hc$age[age_hc$hot_cold == "cold"], na.rm = TRUE)
  cat(sprintf("Median age — hot: %.1f,  cold: %.1f\n", med_hot, med_cold))
  cat(sprintf("Mann-Whitney U: W=%.0f, p=%.2e\n", mw_age$statistic, mw_age$p.value))

  p8b <- ggplot(age_hc, aes(x = hot_cold, y = age, fill = hot_cold)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.9) +
    scale_fill_manual(values = c("hot" = "#e74c3c", "cold" = "#3498db"), guide = "none") +
    labs(
      x     = NULL,
      y     = "Age at diagnosis (years)",
      title = sprintf("Age by immune phenotype (Mann-Whitney p=%.2e)", mw_age$p.value)
    ) +
    theme_bw()

  ggsave(file.path(FIGURES, "age_hot_cold.png"),
         plot = p8b, width = 5, height = 5, dpi = 150)
  p8b
} else {
  cat("age or hot_cold column not found.\n")
}
```

### 8c. Age × senescence Spearman correlations

```{r age-senescence-correlations}
# EXPLORATORY: interpret with BH FDR, not primary threshold
age_senes_cols <- intersect(
  c("CDKN2A_expr", "SASP", "senmayo", "senescence_up", "reactome_senescence",
    "CDKN1A_expr", "SERPINE1_expr"),
  colnames(master)
)

if ("age" %in% colnames(master) && length(age_senes_cols) > 0) {
  age_cor <- master |>
    select(age, all_of(age_senes_cols)) |>
    filter(!is.na(age)) |>
    cor_test(vars  = "age",
             vars2 = all_of(age_senes_cols),
             method = "spearman") |>
    adjust_pvalue(method = "BH") |>
    add_significance() |>
    arrange(desc(abs(cor)))

  print(age_cor)

  # Heatmap (single row)
  age_rho_mat <- age_cor |>
    select(var1, var2, cor) |>
    tidyr::pivot_wider(names_from = var2, values_from = cor) |>
    tibble::column_to_rownames("var1") |>
    as.matrix()

  png(file.path(FIGURES, "age_senescence_correlations.png"),
      width = 900, height = 300, res = 150)
  pheatmap(age_rho_mat,
           display_numbers = TRUE,
           number_format   = "%.2f",
           color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
           breaks           = seq(-0.4, 0.4, length.out = 101),
           main             = "Spearman \u03c1: Patient age vs senescence markers",
           fontsize         = 11,
           cluster_rows     = FALSE)
  dev.off()

  pheatmap(age_rho_mat,
           display_numbers = TRUE,
           number_format   = "%.2f",
           color            = colorRampPalette(c("#2166ac","white","#d6604d"))(100),
           breaks           = seq(-0.4, 0.4, length.out = 101),
           main             = "Spearman \u03c1: Patient age vs senescence markers",
           fontsize         = 11,
           cluster_rows     = FALSE)
} else {
  cat("age column or senescence score columns not found.\n")
}
```

### 8d. Summary table by tumor type

```{r age-summary-table}
if ("age" %in% colnames(master) && "tumor_type" %in% colnames(master)) {
  age_summary <- master |>
    filter(!is.na(tumor_type)) |>
    group_by(tumor_type) |>
    summarise(
      n            = n(),
      median_age   = median(age, na.rm = TRUE),
      pct_hot      = 100 * mean(hot_cold == "hot", na.rm = TRUE),
      median_CDKN2A = median(CDKN2A_expr, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(pct_hot))

  print(age_summary)
} else {
  cat("age or tumor_type column not found.\n")
}
```

## 9. CDKN2A Deletion × Expression 4-Group Analysis

```{r cdkn2a-4group}
# EXPLORATORY: interpret with BH FDR, not primary threshold
if ("cdkn2a_deleted" %in% colnames(master) && "CDKN2A_expr" %in% colnames(master)) {
  # Within-type median of CDKN2A expression
  master <- master |>
    group_by(tumor_type) |>
    mutate(cdkn2a_expr_high = CDKN2A_expr >= median(CDKN2A_expr, na.rm = TRUE)) |>
    ungroup()

  master <- master |>
    mutate(cdkn2a_4group = case_when(
      cdkn2a_deleted & !cdkn2a_expr_high  ~ "del-low",
      cdkn2a_deleted &  cdkn2a_expr_high  ~ "del-high",
     !cdkn2a_deleted & !cdkn2a_expr_high  ~ "intact-low",
     !cdkn2a_deleted &  cdkn2a_expr_high  ~ "intact-high",
      TRUE ~ NA_character_
    ))

  cat("CDKN2A 4-group sizes:\n")
  print(table(master$cdkn2a_4group))

  # Hot fraction by 4-group
  hot_4g <- master |>
    filter(!is.na(cdkn2a_4group), !is.na(hot_cold)) |>
    group_by(cdkn2a_4group) |>
    summarise(
      n            = n(),
      pct_hot      = 100 * mean(hot_cold == "hot", na.rm = TRUE),
      mean_ifng    = mean(ifng_score, na.rm = TRUE),
      mean_SASP    = mean(SASP, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(pct_hot))

  print(hot_4g)

  # Chi-squared: 4-group vs hot/cold
  ct4_data <- master |> filter(!is.na(cdkn2a_4group), !is.na(hot_cold))
  ct4 <- table(group = ct4_data$cdkn2a_4group, immune = ct4_data$hot_cold)
  chi4 <- chisq.test(ct4)
  if (any(chi4$expected < 5))
    warning("Chi-squared (4-group): expected cell counts < 5 — results may be unreliable")
  cat(sprintf("\nChi-squared (4-group vs hot/cold): chi2=%.2f, p=%.2e, df=%d\n",
              chi4$statistic, chi4$p.value, chi4$parameter))

  # Plot: hot fraction by 4-group
  group_colours <- c(
    "del-low"     = "#c0392b",
    "del-high"    = "#e08080",
    "intact-low"  = "#85c1e9",
    "intact-high" = "#1a5276"
  )

  p9 <- ggplot(
    master |> filter(!is.na(cdkn2a_4group), !is.na(hot_cold)),
    aes(x = cdkn2a_4group, fill = hot_cold)
  ) +
    geom_bar(position = "fill") +
    scale_fill_manual(values = c("hot" = "#e74c3c", "cold" = "#3498db")) +
    scale_x_discrete(limits = c("del-low", "del-high", "intact-low", "intact-high")) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      x     = "CDKN2A status",
      y     = "Fraction",
      title = "Immune phenotype by CDKN2A deletion × expression status",
      subtitle = sprintf("Chi-squared p=%.2e", chi4$p.value),
      fill  = NULL
    ) +
    theme_bw()

  ggsave(file.path(FIGURES, "cdkn2a_4group_immune.png"),
         plot = p9, width = 7, height = 5, dpi = 150)
  p9
} else {
  cat("cdkn2a_deleted or CDKN2A_expr not found — re-run notebooks 01–03.\n")
}
```

```{r cdkn2a-purity-quartile}
# ── 9b. CDKN2A expression vs deletion status by purity quartile ──────────────
# Tests whether "high-expression + deleted" samples are purity artefacts
if ("cdkn2a_deleted" %in% names(master) && "tumor_purity" %in% names(master)) {
  pq_df <- master |>
    filter(!is.na(cdkn2a_deleted), !is.na(tumor_purity), !is.na(CDKN2A_expr)) |>
    mutate(
      del_status = ifelse(cdkn2a_deleted, "Deleted", "Intact"),
      purity_q   = ntile(tumor_purity, 4),
      purity_label = paste0("Purity Q", purity_q)
    )

  p_purity_del <- ggplot(pq_df, aes(x = del_status, y = CDKN2A_expr,
                                     fill = del_status)) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
    facet_wrap(~ purity_label, nrow = 1) +
    scale_fill_manual(values = c("Deleted" = "#e74c3c", "Intact" = "#3498db")) +
    labs(
      title = "CDKN2A expression vs deletion status by purity quartile",
      subtitle = "If high-expression+deleted persists in high-purity tumours, it is biological not artefactual",
      x = "", y = "CDKN2A expression (log2)", fill = ""
    ) +
    theme_bw() +
    theme(legend.position = "none", strip.background = element_blank())

  ggsave(file.path(FIGURES, "cdkn2a_expr_vs_del_by_purity.png"),
         p_purity_del, width = 10, height = 4, dpi = 150)
  print(p_purity_del)
} else {
  message("Skipping CDKN2A purity quartile plot: required columns missing.")
}
```

## 10. Thorsson Immune Subtype (C1–C6) × Senescence

```{r immune-subtype-crosstab}
if ("immune_subtype" %in% colnames(master)) {
  cat("Immune subtype distribution:\n")
  print(table(master$immune_subtype, useNA = "ifany"))

  # Cross-tab: immune subtype × hot/cold
  ct_sub <- table(
    subtype = master$immune_subtype[!is.na(master$immune_subtype)],
    immune  = master$hot_cold[!is.na(master$immune_subtype)]
  )
  cat("\nImmune subtype × hot/cold:\n")
  print(ct_sub)
  chi_sub <- chisq.test(ct_sub)
  cat(sprintf("Chi-squared: chi2=%.2f, p=%.2e\n", chi_sub$statistic, chi_sub$p.value))

  # ── 10b. Cramér's V: immune subtype × hot/cold concordance ──────────────────
  if ("immune_subtype" %in% names(master)) {
    ct_tab <- table(master$hot_cold, master$immune_subtype)
    chi2   <- chisq.test(ct_tab)
    n      <- sum(ct_tab)
    k      <- min(nrow(ct_tab), ncol(ct_tab))
    cramers_v <- sqrt(chi2$statistic / (n * (k - 1)))
    message(sprintf("Cramér's V (hot/cold × immune subtype): %.3f  (χ²=%.1f, df=%d, p=%.2e)",
                    cramers_v, chi2$statistic, chi2$parameter, chi2$p.value))

    # Proportion table for interpretation
    prop_tab <- prop.table(ct_tab, margin = 2)  # column proportions
    cat("\nHot fraction within each Thorsson subtype:\n")
    print(round(prop_tab["hot", ] * 100, 1))

    cat(sprintf(
      "\nInterpretation: Cramér's V = %.2f suggests %s concordance between\n",
      cramers_v,
      ifelse(cramers_v > 0.5, "strong", ifelse(cramers_v > 0.3, "moderate", "weak"))
    ))
    cat("hot/cold and Thorsson immune subtypes.\n")
    cat("If V > 0.5, the two classifications are largely redundant.\n")
  }

  # CDKN2A expression by immune subtype
  sub_senes <- master |>
    filter(!is.na(immune_subtype), !is.na(CDKN2A_expr)) |>
    group_by(immune_subtype) |>
    summarise(
      n            = n(),
      median_CDKN2A = median(CDKN2A_expr, na.rm = TRUE),
      median_SASP   = median(SASP, na.rm = TRUE),
      pct_hot       = 100 * mean(hot_cold == "hot", na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(immune_subtype)

  print(sub_senes)

  p10 <- ggplot(
    master |> filter(!is.na(immune_subtype), !is.na(CDKN2A_expr)),
    aes(x = immune_subtype, y = CDKN2A_expr, fill = immune_subtype)
  ) +
    geom_violin(alpha = 0.7, trim = FALSE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
    labs(
      x     = "Thorsson immune subtype",
      y     = "CDKN2A expression (log2)",
      title = "CDKN2A expression by Thorsson immune subtype (C1–C6)"
    ) +
    theme_bw() +
    theme(legend.position = "none")

  ggsave(file.path(FIGURES, "cdkn2a_by_immune_subtype.png"),
         plot = p10, width = 8, height = 5, dpi = 150)
  p10
} else {
  cat("immune_subtype column not found — re-run notebooks 01–02.\n")
}
```

## 11. Stromal Deconfounding of SASP Signal

```{r stromal-deconfounding}
# Test whether SASP → hot/cold association survives purity deconfounding.
# Strategy: regress SASP on tumor_purity, use residuals as "purity-adjusted SASP"
if ("tumor_purity" %in% colnames(master) && "SASP" %in% colnames(master)) {
  sasp_purity_data <- master |>
    filter(!is.na(SASP), !is.na(tumor_purity), !is.na(hot_cold))

  # Correlation: purity vs SASP (stromal inflation check)
  ct_pur <- cor.test(sasp_purity_data$tumor_purity, sasp_purity_data$SASP,
                     method = "spearman", exact = FALSE)
  cat(sprintf("Spearman ρ(purity, SASP): rho=%.3f, p=%.2e\n",
              ct_pur$estimate, ct_pur$p.value))
  cat("  (negative ρ expected: low purity = more stroma = higher SASP)\n\n")

  # Partial correlation: SASP vs IFN-γ score, controlling for purity
  # Implemented as residuals from purity regression
  lm_sasp <- lm(SASP ~ tumor_purity, data = sasp_purity_data)
  lm_ifng <- lm(ifng_score ~ tumor_purity, data = sasp_purity_data)
  resid_sasp <- residuals(lm_sasp)
  resid_ifng <- residuals(lm_ifng)

  ct_partial <- cor.test(resid_sasp, resid_ifng, method = "spearman", exact = FALSE)
  ct_raw     <- cor.test(sasp_purity_data$SASP, sasp_purity_data$ifng_score,
                         method = "spearman", exact = FALSE)

  cat(sprintf("Raw SASP vs IFN-γ:             rho=%.3f, p=%.2e\n",
              ct_raw$estimate, ct_raw$p.value))
  cat(sprintf("Purity-adjusted SASP vs IFN-γ: rho=%.3f, p=%.2e\n",
              ct_partial$estimate, ct_partial$p.value))
  cat("(If adjusted ρ remains significant, SASP signal is not purely stromal.)\n\n")

  # Store purity-adjusted SASP
  master$SASP_purity_adj <- NA_real_
  master$SASP_purity_adj[!is.na(master$SASP) & !is.na(master$tumor_purity)] <- resid_sasp

  # Mann-Whitney: adjusted SASP in hot vs cold
  adj_hot  <- master$SASP_purity_adj[master$hot_cold == "hot"  & !is.na(master$SASP_purity_adj)]
  adj_cold <- master$SASP_purity_adj[master$hot_cold == "cold" & !is.na(master$SASP_purity_adj)]
  mw_adj <- wilcox.test(adj_hot, adj_cold, exact = FALSE)
  cat(sprintf("Mann-Whitney (purity-adjusted SASP, hot vs cold): p=%.2e\n", mw_adj$p.value))
  cat(sprintf("  Median adj SASP — hot: %.3f, cold: %.3f\n",
              median(adj_hot, na.rm=TRUE), median(adj_cold, na.rm=TRUE)))

} else {
  cat("tumor_purity or SASP column not found — re-run notebooks 01–03.\n")
}

# ── 11b. Purity-adjusted IFN-γ score ─────────────────────────────────────────
if ("tumor_purity" %in% names(master) && "ifng_score" %in% names(master)) {
  purity_df <- master |> filter(!is.na(tumor_purity), !is.na(ifng_score))
  # Residualise ifng_score on tumor_purity
  lm_ifng <- lm(ifng_score ~ tumor_purity, data = purity_df)
  purity_df <- purity_df |> mutate(ifng_resid = residuals(lm_ifng))

  rho_raw  <- cor(purity_df$tumor_purity, purity_df$ifng_score, method = "spearman")
  rho_adj  <- cor(purity_df$tumor_purity, purity_df$ifng_resid, method = "spearman")
  message(sprintf("IFN-γ ~ purity: raw ρ = %.3f; after residualisation ρ ≈ %.3f", rho_raw, rho_adj))

  # Mann-Whitney: hot vs cold using purity-adjusted IFN-γ
  wt_ifng <- wilcox.test(ifng_resid ~ hot_cold, data = purity_df)
  message(sprintf("Hot vs cold purity-adj IFN-γ: W = %.0f, p = %.3e", wt_ifng$statistic, wt_ifng$p.value))
  cat("\nNote: Both SASP and IFN-γ scores are tested with and without purity adjustment above.\n")
} else {
  message("Skipping IFN-γ purity adjustment: tumor_purity or ifng_score not available.")
}
```

## 12. Reproducibility

```{r session-info}
sessionInfo()
```
