---
title: "05 — Survival Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

Tests whether the combination of hot/cold immune phenotype and p16/CDKN2A status
predicts patient overall survival in TCGA.

**Analyses:**

1. Kaplan-Meier: hot vs cold
2. Kaplan-Meier: 4-group (hot/cold × p16-high/low)
3. Sensitivity: per-tumor-type log-rank
4. Cox proportional hazards: adjusted for tumor type via strata

```{r setup, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(broom)
library(jsonlite)
library(cmprsk)

set.seed(42)

DATA    <- "../data"
FIGURES <- normalizePath("../figures", mustWork = FALSE)
dir.create(FIGURES, showWarnings = FALSE)
```

## 1. Load & Merge Data

```{r load-data}
stopifnot(
  file.exists(file.path(DATA, "sample_immune_labels.parquet")),
  file.exists(file.path(DATA, "sample_senescence.parquet")),
  file.exists(file.path(DATA, "tcga_clinical_survival.tsv"))
)

immune <- read_parquet(file.path(DATA, "sample_immune_labels.parquet")) |>
  rename(sample_id = `__index_level_0__`)
senes  <- read_parquet(file.path(DATA, "sample_senescence.parquet")) |>
  rename(sample_id = `__index_level_0__`)

# Clinical / survival data (Xena format)
# Common survival columns: OS (1=deceased/0=censored), OS.time (days)
clin <- read.table(
  file.path(DATA, "tcga_clinical_survival.tsv"),
  sep = "\t", header = TRUE, row.names = 1, stringsAsFactors = FALSE
)
cat("Clinical columns:", paste(colnames(clin), collapse = ", "), "\n")

OS_COL  <- "OS"
OST_COL <- "OS.time"

# Include DSS and PFI endpoints alongside OS; take only columns that exist
extra_clin_cols <- intersect(
  c(OS_COL, OST_COL, "DSS", "DSS.time", "PFI", "PFI.time",
    "age_at_initial_pathologic_diagnosis"),
  colnames(clin)
)

master <- immune |>
  inner_join(senes, by = "sample_id") |>
  inner_join(
    clin[, extra_clin_cols, drop = FALSE] |>
      tibble::rownames_to_column("sample_id"),
    by = "sample_id"
  ) |>
  rename(age = age_at_initial_pathologic_diagnosis) |>
  filter(!is.na(.data[[OS_COL]]), !is.na(.data[[OST_COL]])) |>
  mutate(
    OS.time_years  = .data[[OST_COL]] / 365.25,
    DSS.time_years = if ("DSS.time" %in% names(.)) DSS.time / 365.25 else NA_real_,
    PFI.time_years = if ("PFI.time" %in% names(.)) PFI.time / 365.25 else NA_real_
  )

cat("Survival table:", nrow(master), "rows\n")
cat("Endpoints available:",
    paste(intersect(c("OS","DSS","PFI"), colnames(master)), collapse=", "), "\n")
```

## 1b. Join Tumor Purity and TMB

```{r load-tmb-purity}
# ── Tumor purity ──────────────────────────────────────────────────────────────
purity_file <- file.path(DATA, "tcga_purity.tsv")
if (file.exists(purity_file)) {
  purity_df <- read.table(purity_file, sep = "\t", header = TRUE,
                          stringsAsFactors = FALSE)
  samp_col <- grep("sample|barcode", colnames(purity_df), ignore.case = TRUE, value = TRUE)[1]
  pur_col  <- grep("purity|absolute", colnames(purity_df), ignore.case = TRUE, value = TRUE)[1]
  if (!is.na(samp_col) && !is.na(pur_col)) {
    pur_tbl <- purity_df[, c(samp_col, pur_col)]
    colnames(pur_tbl) <- c("sample_id", "tumor_purity")
    pur_tbl$sample_id  <- gsub("\\.", "-", pur_tbl$sample_id)
    pur_tbl$sample_id  <- sub("^(TCGA-[^-]+-[^-]+-[0-9]{2}).*", "\\1", pur_tbl$sample_id)
    pur_tbl$tumor_purity <- suppressWarnings(as.numeric(pur_tbl$tumor_purity))
    master <- left_join(master, pur_tbl, by = "sample_id")
    cat(sprintf("Purity joined: %d / %d samples (median=%.2f)\n",
                sum(!is.na(master$tumor_purity)), nrow(master),
                median(master$tumor_purity, na.rm = TRUE)))
  } else {
    cat("Could not identify purity/sample columns\n")
    master$tumor_purity <- NA_real_
  }
} else {
  cat("tcga_purity.tsv not found — run notebook 01 first\n")
  master$tumor_purity <- NA_real_
}

# ── TMB ───────────────────────────────────────────────────────────────────────
tmb_file <- file.path(DATA, "tcga_tmb.json")
if (file.exists(tmb_file)) {
  tmb_raw <- tryCatch(fromJSON(tmb_file), error = function(e) NULL)
  if (!is.null(tmb_raw) && is.data.frame(tmb_raw) &&
      all(c("sampleId", "value") %in% colnames(tmb_raw))) {
    tmb_tbl <- data.frame(
      sample_id = tmb_raw$sampleId,
      tmb       = suppressWarnings(as.numeric(tmb_raw$value)),
      stringsAsFactors = FALSE
    )
    master <- left_join(master, tmb_tbl, by = "sample_id")
    cat(sprintf("TMB joined: %d / %d samples (median=%.1f mut/Mb)\n",
                sum(!is.na(master$tmb)), nrow(master),
                median(master$tmb, na.rm = TRUE)))
  } else {
    cat("TMB file empty or unexpected format\n")
    master$tmb <- NA_real_
  }
} else {
  cat("tcga_tmb.json not found — run notebook 01 first\n")
  master$tmb <- NA_real_
}
```

## 2. Classify p16-High/Low Within Tumor Type

```{r p16-classify}
# ntile(2) assigns 1=low, 2=high within each tumor type
master <- master |>
  group_by(tumor_type) |>
  mutate(
    p16_status = case_when(
      ntile(CDKN2A_expr, 2) == 2 ~ "p16-high",
      ntile(CDKN2A_expr, 2) == 1 ~ "p16-low",
      TRUE ~ NA_character_
    )
  ) |>
  ungroup()

# Minimum 2 p16-status levels per tumor type required for within-type split
# Remove tumor types where ntile produced only one level (n too small or tied)
master <- master |>
  group_by(tumor_type) |>
  filter(n_distinct(p16_status, na.rm = TRUE) == 2) |>
  ungroup()

master <- master |>
  mutate(
    group4 = paste(hot_cold, p16_status, sep = " / ")
  )

cat("Group sizes:\n")
print(table(master$group4))
```

## 3. KM: Hot vs Cold

```{r km-hot-cold}
cat(sprintf("Events: %d / %d samples; median OS: %.1f yrs\n",
            sum(master$OS), nrow(master), median(master$OS.time_years)))

# NOTE: The KM curves below are UNADJUSTED for age and tumor purity.
# Age is included in the Cox model (age HR, see above).
# If hot tumors are systematically younger than cold tumors (see notebook 04 §8b),
# part of the survival advantage may reflect age rather than immune phenotype.
# Interpretation: KM shows the marginal association; Cox HR is the age-adjusted estimate.

surv_obj <- Surv(master$OS.time_years, master$OS)

fit_hc <- survfit(surv_obj ~ hot_cold, data = master)

# Log-rank test
lr_hc <- survdiff(surv_obj ~ hot_cold, data = master)
lr_p  <- 1 - pchisq(lr_hc$chisq, df = length(lr_hc$obs) - 1)
cat(sprintf("Log-rank p (hot vs cold): %.2e\n", lr_p))

p1 <- ggsurvplot(
  fit_hc,
  data       = master,
  palette    = c("cold" = "#3498db", "hot" = "#e74c3c"),
  pval       = TRUE,
  conf.int   = TRUE,
  risk.table = TRUE,
  xlab       = "Time (years)",
  title      = sprintf("Overall Survival: Hot vs Cold (log-rank p=%.2e)", lr_p),
  legend.labs = c("cold", "hot")
)

ggsave(file.path(FIGURES, "km_hot_cold.png"),
       plot = p1, width = 8, height = 8, dpi = 150)
p1
```

```{r tertile-sensitivity, fig.width=10, fig.height=7}
# ── Sensitivity: tertile split for hot/cold IFN-γ threshold ──────────────────
# Tests whether survival conclusions depend on the choice of median vs tertile cut

master_tertile <- master |>
  mutate(
    ifng_tertile = ntile(ifng_score, 3),
    hot_cold_tertile = case_when(
      ifng_tertile == 1 ~ "cold (bottom tertile)",
      ifng_tertile == 3 ~ "hot (top tertile)",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(hot_cold_tertile))

km_tertile <- survfit(
  Surv(OS.time_years, OS) ~ hot_cold_tertile,
  data = master_tertile
)
lr_tertile <- survdiff(
  Surv(OS.time_years, OS) ~ hot_cold_tertile,
  data = master_tertile
)
p_tertile <- 1 - pchisq(lr_tertile$chisq, df = length(lr_tertile$obs) - 1)
message(sprintf("Tertile split log-rank p = %.4f (n = %d, excludes middle tertile)",
                p_tertile, nrow(master_tertile)))

ggsurvplot(
  km_tertile, data = master_tertile,
  palette      = c("#e74c3c", "#3498db"),
  pval         = TRUE,
  risk.table   = TRUE,
  risk.table.height = 0.28,
  xlab         = "Time (years)",
  ylab         = "Overall survival",
  title        = "Sensitivity: OS by hot/cold (tertile cut, median excluded)",
  legend.labs  = c("hot (top tertile)", "cold (bottom tertile)")
)
```

## 4. KM: 4-Group (Hot/Cold × p16-High/Low)

```{r km-4group}
sub4 <- master |>
  filter(!is.na(group4), !is.na(OS.time_years), !is.na(OS)) |>
  group_by(group4) |>
  filter(n() >= 30) |>
  ungroup()

# Filter out group4 combinations with <5 events (too sparse for reliable KM)
sub4 <- sub4 |>
  group_by(group4) |>
  filter(sum(OS, na.rm = TRUE) >= 5) |>
  ungroup()

surv4 <- Surv(sub4$OS.time_years, sub4$OS)
fit4  <- survfit(surv4 ~ group4, data = sub4)

lr4   <- survdiff(surv4 ~ group4, data = sub4)
lr4_p <- 1 - pchisq(lr4$chisq, df = length(lr4$obs) - 1)
cat(sprintf("Log-rank p (4-group): %.2e\n", lr4_p))

group_palette <- c(
  "hot / p16-high"  = "#c0392b",
  "hot / p16-low"   = "#e08080",
  "cold / p16-high" = "#1a5276",
  "cold / p16-low"  = "#85c1e9"
)

p2 <- ggsurvplot(
  fit4,
  data        = sub4,
  palette     = group_palette[levels(factor(sub4$group4))],
  pval        = TRUE,
  conf.int    = FALSE,
  risk.table  = TRUE,
  xlab        = "Time (years)",
  title       = sprintf("OS: Hot/Cold \u00d7 p16-High/Low (log-rank p=%.2e)", lr4_p),
  legend.labs = levels(factor(sub4$group4)),
  fontsize    = 3.5
)

ggsave(file.path(FIGURES, "km_4group.png"),
       plot = p2, width = 9, height = 9, dpi = 150)
p2
```

## 5. Sensitivity: Per-Tumor-Type Log-Rank

```{r per-type-logrank}
per_type_lr <- master |>
  filter(!is.na(hot_cold), !is.na(OS.time_years), !is.na(OS)) |>
  group_by(tumor_type) |>
  filter(n() >= 50) |>
  summarise(
    n_hot    = sum(hot_cold == "hot"),
    n_cold   = sum(hot_cold == "cold"),
    lr_pval  = {
      sd <- survdiff(Surv(OS.time_years, OS) ~ hot_cold,
                     data = cur_data())
      1 - pchisq(sd$chisq, df = 1)
    },
    .groups = "drop"
  ) |>
  mutate(padj = p.adjust(lr_pval, method = "BH")) |>
  arrange(lr_pval)

print(per_type_lr)
```

## 6. Cox Proportional Hazards (Adjusted)

```{r cox-ph}
# Include new senescence expression markers where available
new_expr_markers <- intersect(
  c("CDKN1A_expr", "SERPINE1_expr", "GDF15_expr"),
  colnames(master)
)
# Optional covariates: include tumor_purity and TMB when available
optional_covars <- intersect(c("tumor_purity", "tmb"), colnames(master))
# Log-transform TMB (heavy right skew)
if ("tmb" %in% colnames(master)) master$log_tmb <- log1p(master$tmb)
optional_covars <- c(setdiff(optional_covars, "tmb"),
                     if ("tmb" %in% optional_covars) "log_tmb")

cox_vars <- c("OS.time_years", "OS", "age", "ifng_score", "CDKN2A_expr",
              new_expr_markers, optional_covars, "tumor_type")

# Filter only on required covariates; optional covariates (purity, TMB) handled by na.action=na.omit
required_cox_vars <- c("OS.time_years", "OS", "age", "ifng_score", "CDKN2A_expr", "tumor_type")
cox_data <- master |>
  select(all_of(intersect(cox_vars, colnames(master)))) |>
  filter(if_all(all_of(intersect(required_cox_vars, colnames(master))), Negate(is.na)))

cat(sprintf("Cox data: %d samples after NA removal (%.1f%% of master)\n",
            nrow(cox_data), 100 * nrow(cox_data) / nrow(master)))

# Build formula: age-, purity-, and TMB-adjusted where covariates are available
cox_rhs <- paste(
  c("age", "ifng_score", "CDKN2A_expr", new_expr_markers,
    optional_covars, "strata(tumor_type)"),
  collapse = " + "
)
cox_formula <- as.formula(paste("Surv(OS.time_years, OS) ~", cox_rhs))
cat("Cox formula:", deparse(cox_formula), "\n")

# strata(tumor_type) allows baseline hazard to vary by cancer type
# without fitting one coefficient per type (avoids one-hot explosion)
cox_fit <- coxph(
  cox_formula,
  data = cox_data
)

print(summary(cox_fit))

# Tidy output
cox_tidy <- tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE)
print(cox_tidy)

# Multicollinearity check (VIF > 5 suggests problematic collinearity)
vif_res <- tryCatch(car::vif(cox_fit), error = function(e) NULL)
if (!is.null(vif_res)) {
  cat("\nVariance Inflation Factors (VIF > 5 suggests multicollinearity):\n")
  print(round(vif_res, 2))
}

# Check proportional hazards assumption
ph_test <- cox.zph(cox_fit)
print(ph_test)
cat("\nPH assumption satisfied if all p > 0.05 (non-significant deviation)\n")

# Identify and report PH-violating terms; run landmark sensitivity analysis if any found
ph_df <- as.data.frame(ph_test$table)
ph_df$term <- rownames(ph_df)
violators <- ph_df[!is.na(ph_df$p) & ph_df$p < 0.05 & ph_df$term != "GLOBAL", ]
if (nrow(violators) > 0) {
  cat(sprintf("\nPH VIOLATION: %d term(s) violate proportional hazards (p<0.05):\n",
              nrow(violators)))
  print(violators[, c("term", "p")])
  cat("Landmark sensitivity analysis at median follow-up:\n")
  t_med <- median(cox_data$OS.time_years, na.rm = TRUE)
  lm_data <- cox_data[cox_data$OS.time_years >= t_med, ]
  lm_data$OS.time_years <- lm_data$OS.time_years - t_med
  lm_formula <- as.formula(paste("Surv(OS.time_years, OS) ~", cox_rhs))
  lm_fit <- tryCatch(coxph(lm_formula, data = lm_data), error = function(e) NULL)
  if (!is.null(lm_fit)) {
    cat(sprintf("  (n=%d, follow-up > %.2f yr)\n", nrow(lm_data), t_med))
    print(tidy(lm_fit, exponentiate = TRUE, conf.int = TRUE)[,
          c("term", "estimate", "conf.low", "conf.high", "p.value")])
  }
} else {
  cat("All terms satisfy PH assumption.\n")
}

# Forest plot — ggforest doesn't support strata(), so build manually from tidy output
p3 <- ggplot(cox_tidy, aes(x = estimate, xmin = conf.low, xmax = conf.high, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(height = 0.2) +
  geom_point(size = 3, colour = "#e74c3c") +
  scale_x_log10() +
  labs(
    x     = "Hazard Ratio (95% CI)",
    y     = NULL,
    title = "Cox HR (adjusted for tumor type via strata)"
  ) +
  theme_bw()

ggsave(file.path(FIGURES, "cox_forest_plot.png"),
       plot = p3, width = 8, height = 4, dpi = 150)
p3

# ── Interaction Cox: CDKN2A_expr × ifng_score ─────────────────────────────────
# Tests whether the senescence × immune phenotype synergy predicts survival
# above and beyond their additive effects.
if (all(c("CDKN2A_expr", "ifng_score") %in% names(cox_data))) {
  cox_interaction_formula <- as.formula(paste(
    "Surv(OS.time_years, OS) ~ age + CDKN2A_expr * ifng_score +",
    paste(c(intersect(optional_covars, names(cox_data)), "strata(tumor_type)"), collapse = " + ")
  ))

  cox_int <- tryCatch(
    coxph(cox_interaction_formula, data = cox_data),
    error = function(e) { warning("Interaction Cox failed: ", e$message); NULL }
  )

  if (!is.null(cox_int)) {
    int_tidy <- broom::tidy(cox_int, conf.int = TRUE, exponentiate = TRUE)
    int_row  <- int_tidy |> filter(grepl(":", term))
    if (nrow(int_row) > 0) {
      message(sprintf(
        "CDKN2A_expr × ifng_score interaction: HR = %.3f [%.3f–%.3f], p = %.3e",
        int_row$estimate, int_row$conf.low, int_row$conf.high, int_row$p.value
      ))
      cat("\nInteraction term (CDKN2A × IFN-γ):\n")
      print(int_row |> select(term, estimate, conf.low, conf.high, p.value))
    }
    # LRT: additive (main Cox) vs interaction model
    lrt_res <- tryCatch(anova(cox_fit, cox_int), error = function(e) NULL)
    if (!is.null(lrt_res) && nrow(lrt_res) >= 2) {
      message(sprintf(
        "LRT (additive vs interaction model): chi2=%.2f, df=%d, p=%.3e",
        lrt_res$Chisq[2], lrt_res$Df[2], lrt_res[2, "Pr(>|Chi|)"]
      ))
    }

    # Export interaction results alongside main Cox results
    write.csv(
      int_tidy,
      file.path(DATA, "cox_interaction_results.csv"),
      row.names = FALSE
    )
    message("Saved cox_interaction_results.csv")
  }
} else {
  message("Skipping interaction Cox: CDKN2A_expr or ifng_score not in dataset.")
}
```

```{r log-log-ph-check, fig.width=10, fig.height=5}
# Log-log survival plot for visual PH assumption check
# Parallel lines → PH holds; converging/diverging → violation

km_hc <- survfit(Surv(OS.time_years, OS) ~ hot_cold, data = master)
plot(km_hc, fun = "cloglog",
     col  = c("#e74c3c", "#3498db"),
     xlab = "log(Time, years)",
     ylab = "log(-log(S(t)))",
     main = "Log-log plot: hot vs cold (parallel = PH holds)",
     lwd  = 2)
legend("topleft", legend = c("hot", "cold"),
       col = c("#e74c3c", "#3498db"), lwd = 2, bty = "n")
```

```{r log-log-p16, fig.width=10, fig.height=5}
km_p16 <- survfit(Surv(OS.time_years, OS) ~ p16_status, data = master)
plot(km_p16, fun = "cloglog",
     col  = c("#27ae60", "#e67e22"),
     xlab = "log(Time, years)",
     ylab = "log(-log(S(t)))",
     main = "Log-log plot: p16 high vs low",
     lwd  = 2)
legend("topleft", legend = c("p16-high", "p16-low"),
       col = c("#27ae60", "#e67e22"), lwd = 2, bty = "n")
```

## 7. KM: TP53 Mutation Status

```{r km-tp53}
if ("tp53_mutated" %in% colnames(master)) {
  tp53_data <- master |>
    filter(!is.na(tp53_mutated), !is.na(OS.time_years), !is.na(OS)) |>
    mutate(tp53_status = ifelse(tp53_mutated, "TP53 mutated", "TP53 wild-type"))

  surv_tp53 <- Surv(tp53_data$OS.time_years, tp53_data$OS)
  fit_tp53  <- survfit(surv_tp53 ~ tp53_status, data = tp53_data)

  lr_tp53   <- survdiff(surv_tp53 ~ tp53_status, data = tp53_data)
  lr_tp53_p <- 1 - pchisq(lr_tp53$chisq, df = length(lr_tp53$obs) - 1)
  cat(sprintf("TP53 mutation frequency: %.1f%%\n",
              100 * mean(tp53_data$tp53_mutated, na.rm = TRUE)))
  cat(sprintf("Log-rank p (TP53 mutated vs WT): %.2e\n", lr_tp53_p))

  p_tp53 <- ggsurvplot(
    fit_tp53,
    data        = tp53_data,
    palette     = c("TP53 mutated" = "#8e44ad", "TP53 wild-type" = "#95a5a6"),
    pval        = TRUE,
    conf.int    = TRUE,
    risk.table  = TRUE,
    xlab        = "Time (years)",
    title       = sprintf("Overall Survival: TP53 Mutated vs WT (log-rank p=%.2e)", lr_tp53_p),
    legend.labs = c("TP53 mutated", "TP53 wild-type")
  )

  ggsave(file.path(FIGURES, "km_tp53_mutation.png"),
         plot = p_tp53, width = 8, height = 8, dpi = 150)
  p_tp53
} else {
  cat("tp53_mutated column not found — re-run notebooks 01 and 03 first.\n")
}
```

## 8. Alternative Endpoints: DSS and PFI

```{r dss-endpoint}
# Disease-Specific Survival: event = death from cancer (DSS==1)
# Censored includes alive + non-cancer deaths — cleanest test of cancer biology
if ("DSS" %in% colnames(master) && "DSS.time_years" %in% colnames(master)) {
  dss_data <- master |>
    filter(!is.na(DSS), !is.na(DSS.time_years), !is.na(hot_cold))

  cat(sprintf("DSS: %d samples, %d events (cancer deaths)\n",
              nrow(dss_data), sum(dss_data$DSS, na.rm = TRUE)))

  surv_dss <- Surv(dss_data$DSS.time_years, dss_data$DSS)
  fit_dss  <- survfit(surv_dss ~ hot_cold, data = dss_data)
  lr_dss   <- survdiff(surv_dss ~ hot_cold, data = dss_data)
  lr_dss_p <- 1 - pchisq(lr_dss$chisq, df = length(lr_dss$obs) - 1)
  cat(sprintf("DSS log-rank (hot vs cold): p=%.2e\n", lr_dss_p))

  p_dss <- ggsurvplot(
    fit_dss, data = dss_data,
    palette     = c("cold" = "#3498db", "hot" = "#e74c3c"),
    pval        = TRUE, conf.int = TRUE, risk.table = TRUE,
    xlab        = "Time (years)",
    title       = sprintf("Disease-Specific Survival: Hot vs Cold (p=%.2e)", lr_dss_p),
    legend.labs = c("cold", "hot")
  )
  ggsave(file.path(FIGURES, "km_dss_hot_cold.png"),
         plot = p_dss, width = 8, height = 8, dpi = 150)
  p_dss
} else {
  cat("DSS columns not found in clinical data — check tcga_clinical_survival.tsv.\n")
}
```

```{r pfi-endpoint}
# Progression-Free Interval: event = recurrence or progression
# Tests whether senescence/immune phenotype affects recurrence, not just death
if ("PFI" %in% colnames(master) && "PFI.time_years" %in% colnames(master)) {
  pfi_data <- master |>
    filter(!is.na(PFI), !is.na(PFI.time_years), !is.na(hot_cold))

  cat(sprintf("PFI: %d samples, %d events (progressions)\n",
              nrow(pfi_data), sum(pfi_data$PFI, na.rm = TRUE)))

  surv_pfi <- Surv(pfi_data$PFI.time_years, pfi_data$PFI)
  fit_pfi  <- survfit(surv_pfi ~ hot_cold, data = pfi_data)
  lr_pfi   <- survdiff(surv_pfi ~ hot_cold, data = pfi_data)
  lr_pfi_p <- 1 - pchisq(lr_pfi$chisq, df = length(lr_pfi$obs) - 1)
  cat(sprintf("PFI log-rank (hot vs cold): p=%.2e\n", lr_pfi_p))

  p_pfi <- ggsurvplot(
    fit_pfi, data = pfi_data,
    palette     = c("cold" = "#3498db", "hot" = "#e74c3c"),
    pval        = TRUE, conf.int = TRUE, risk.table = TRUE,
    xlab        = "Time (years)",
    title       = sprintf("Progression-Free Interval: Hot vs Cold (p=%.2e)", lr_pfi_p),
    legend.labs = c("cold", "hot")
  )
  ggsave(file.path(FIGURES, "km_pfi_hot_cold.png"),
         plot = p_pfi, width = 8, height = 8, dpi = 150)
  p_pfi
} else {
  cat("PFI columns not found in clinical data.\n")
}
```

## 8b. Competing Risks (Fine-Gray)

```{r competing-risks}
# Fine-Gray model: cancer death (DSS==1) as event; non-cancer death as competing risk
# Event coding: 1 = cancer death, 2 = competing (non-cancer death), 0 = censored
if ("DSS" %in% colnames(master) && "OS" %in% colnames(master) &&
    "DSS.time_years" %in% colnames(master)) {

  cr_data <- master |>
    filter(!is.na(DSS), !is.na(OS), !is.na(DSS.time_years),
           !is.na(ifng_score), !is.na(CDKN2A_expr), !is.na(age), !is.na(tumor_type)) |>
    mutate(
      cr_status = case_when(
        DSS == 1             ~ 1L,  # cancer death (event of interest)
        OS == 1 & DSS == 0  ~ 2L,  # non-cancer death (competing event)
        TRUE                ~ 0L   # censored
      )
    )

  cat(sprintf("Competing risks: %d samples\n", nrow(cr_data)))
  cat("Event counts (0=censored, 1=cancer death, 2=competing):\n")
  print(table(cr_data$cr_status))

  # Covariate matrix (crr requires numeric matrix, no strata support)
  cr_cov_cols <- intersect(c("age", "ifng_score", "CDKN2A_expr",
                              new_expr_markers, optional_covars),
                           colnames(cr_data))
  cov_mat <- as.matrix(cr_data[, cr_cov_cols])

  fg_fit <- tryCatch(
    crr(ftime    = cr_data$DSS.time_years,
        fstatus  = cr_data$cr_status,
        cov1     = cov_mat,
        failcode = 1L,
        cencode  = 0L),
    error = function(e) {
      warning(sprintf("Fine-Gray crr() failed: %s\nCheck event counts and covariate matrix for collinearity.", e$message))
      NULL
    }
  )

  if (is.null(fg_fit)) {
    message("Competing risks model did not converge — skipping forest plot and CSV export for this model.")
  } else {
    cr_sum <- summary(fg_fit)
    cr_tidy <- data.frame(
      term      = rownames(cr_sum$coef),
      HR        = exp(cr_sum$coef[, "coef"]),
      conf.low  = exp(cr_sum$coef[, "coef"] - 1.96 * cr_sum$coef[, "se(coef)"]),
      conf.high = exp(cr_sum$coef[, "coef"] + 1.96 * cr_sum$coef[, "se(coef)"]),
      p.value   = cr_sum$coef[, "p-value"],
      stringsAsFactors = FALSE
    )
    cat("\nFine-Gray sub-distribution HRs (event = cancer death):\n")
    print(cr_tidy)

    # Forest plot
    p_cr <- ggplot(cr_tidy,
                   aes(x = HR, xmin = conf.low, xmax = conf.high, y = term)) +
      geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
      geom_errorbarh(height = 0.2) +
      geom_point(size = 3, colour = "#8e44ad") +
      scale_x_log10() +
      labs(x = "Sub-distribution HR (95% CI)",
           y = NULL,
           title = "Fine-Gray competing risks (event = cancer-specific death)") +
      theme_bw()

    ggsave(file.path(FIGURES, "competing_risks_forest.png"),
           plot = p_cr, width = 8, height = 4, dpi = 150)
    p_cr

    write.csv(cr_tidy, file.path(DATA, "competing_risks_results.csv"), row.names = FALSE)
    cat("Exported competing_risks_results.csv\n")
  }
} else {
  cat("DSS/OS columns not found — competing risks analysis skipped.\n")
}
```

## 9. Age-Stratified KM

```{r km-age-groups}
if ("age" %in% colnames(master)) {
  # Split at median age within each tumor type to avoid cancer-type confounding
  age_km_data <- master |>
    filter(!is.na(age), !is.na(OS.time_years), !is.na(OS)) |>
    group_by(tumor_type) |>
    mutate(age_group = ifelse(age >= median(age, na.rm = TRUE),
                              "older-half", "younger-half")) |>
    ungroup()

  surv_age <- Surv(age_km_data$OS.time_years, age_km_data$OS)
  fit_age  <- survfit(surv_age ~ age_group, data = age_km_data)

  lr_age   <- survdiff(surv_age ~ age_group, data = age_km_data)
  lr_age_p <- 1 - pchisq(lr_age$chisq, df = length(lr_age$obs) - 1)
  cat(sprintf("Log-rank p (older vs younger half): %.2e\n", lr_age_p))

  p_age <- ggsurvplot(
    fit_age,
    data        = age_km_data,
    palette     = c("older-half" = "#8e44ad", "younger-half" = "#27ae60"),
    pval        = TRUE,
    conf.int    = TRUE,
    risk.table  = TRUE,
    xlab        = "Time (years)",
    title       = sprintf("OS by age group (within-type median split; log-rank p=%.2e)", lr_age_p),
    legend.labs = c("older-half", "younger-half")
  )

  ggsave(file.path(FIGURES, "km_age_groups.png"),
         plot = p_age, width = 8, height = 8, dpi = 150)
  p_age
} else {
  cat("age column not found — re-check clinical data join.\n")
}
```

## 10. Export Tidy Results

```{r export-results}
write.csv(cox_tidy,    file.path(DATA, "cox_results.csv"),      row.names = FALSE)
write.csv(per_type_lr, file.path(DATA, "logrank_per_type.csv"), row.names = FALSE)

# DSS log-rank per tumor type (mirrors OS analysis)
if ("DSS" %in% colnames(master) && "DSS.time_years" %in% colnames(master)) {
  per_type_lr_dss <- master |>
    filter(!is.na(hot_cold), !is.na(DSS.time_years), !is.na(DSS)) |>
    group_by(tumor_type) |>
    filter(n() >= 50) |>
    summarise(
      lr_pval_dss = {
        sd <- survdiff(Surv(DSS.time_years, DSS) ~ hot_cold, data = cur_data())
        1 - pchisq(sd$chisq, df = 1)
      },
      .groups = "drop"
    ) |>
    mutate(padj_dss = p.adjust(lr_pval_dss, method = "BH"))

  per_type_lr <- left_join(per_type_lr, per_type_lr_dss, by = "tumor_type")
  write.csv(per_type_lr, file.path(DATA, "logrank_per_type.csv"), row.names = FALSE)
}

cat("Exported cox_results.csv and logrank_per_type.csv\n")
```

## 11. Reproducibility

```{r session-info}
sessionInfo()
```
